---
title: "Final_project_script"
output: html_document
date: "2026-04-16"
---

# ----------LOAD THE PACKAGES FOR PROJECT------------
```{r}

#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(lubridate)
library(here)
library(vip)          
library(xgboost)      
library(ranger)       
library(tidyverse)    
library(doParallel)   
library(shapviz)
library(tidymodels)
library(recipes)
library(rsample)

set.seed(19823)

```

# ----------LOAD TRAINING DATA------------
```{r}
#| message: false
#| warning: false

train_meta <- read_csv("../data/training/training_meta.csv") %>%
  clean_names()

train_trait <- read_csv("../data/training/training_trait.csv") %>%
  clean_names()

train_soil <- read_csv("../data/training/training_soil.csv") %>%
  clean_names()

```

# ------------AGGREGATE TRAINING TRAITS-------------
```{r}
#| message: false
#| warning: false

train_trait_clean <- train_trait %>%
  mutate(
    date_planted = mdy(date_planted),
    date_harvested = mdy(date_harvested)
  ) %>%
  group_by(site, year, hybrid) %>%
  summarize(
    yield = mean(yield_mg_ha, na.rm = TRUE),
    grain_moisture = mean(grain_moisture, na.rm = TRUE),
    date_planted = suppressWarnings(min(date_planted, na.rm = TRUE)),
    date_harvested = suppressWarnings(max(date_harvested, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    date_planted = if_else(is.infinite(date_planted), as.Date(NA), date_planted),
    date_harvested = if_else(is.infinite(date_harvested), as.Date(NA), date_harvested)
  )

head(train_trait_clean)
```

# -------------MERGE------------
```{r}
#| message: false
#| warning: false

train_full <- train_trait_clean %>%
  left_join(train_meta, by = c("site", "year")) %>%
  left_join(train_soil, by = c("site", "year")) %>%
  mutate(across(c(soilp_h, om_pct, soilk_ppm, soilp_ppm), as.numeric))

glimpse(train_full)

```

# ----------LOAD TESTING DATA------------
```{r}
#| message: false
#| warning: false

test_meta <- read_csv("../data/testing/testing_meta.csv") %>%
  clean_names()

test_soil <- read_csv("../data/testing/testing_soil.csv") %>%
  clean_names()

test_sub <- read_csv("../data/testing/testing_submission.csv") %>%
  clean_names()

head(test_sub)
head(test_meta)
head(test_soil)

table(test_meta$previous_crop)
summary(test_soil)
summary(test_sub)
```

# ----------FIXING SOIL SITE COLUMN---------------
```{r}
#| message: false
#| warning: false

test_soil <- read_csv("../data/testing/testing_soil.csv", show_col_types = FALSE) %>%
  clean_names() %>%
  mutate(
    site = sub("_\\d{4}$", "", site),
    year = as.integer(year)
  ) %>%
  group_by(site, year) %>%
  summarize(
    soilp_h = mean(soilp_h, na.rm = TRUE),
    om_pct = mean(om_pct, na.rm = TRUE),
    soilk_ppm = mean(soilk_ppm, na.rm = TRUE),
    soilp_ppm = mean(soilp_ppm, na.rm = TRUE),
    .groups = "drop"
  )

head(test_soil)
```

# -----------CLEAN META--------------
```{r}
#| message: false
#| warning: false

test_meta <- read_csv("../data/testing/testing_meta.csv", show_col_types = FALSE) %>%
  clean_names() %>%
  mutate(
    previous_crop = str_to_lower(previous_crop),
    year = as.integer(year)
  )
```

# -------------MERGE------------
```{r}
#| message: false
#| warning: false

test_sub <- read_csv("../data/testing/testing_submission.csv", show_col_types = FALSE) %>%
  clean_names() %>%
  rename(yield = yield_mg_ha) %>%
  mutate(year = as.integer(year))

test_full <- test_sub %>%
  left_join(test_meta, by = c("site", "year")) %>%
  left_join(test_soil, by = c("site", "year"))
```

#----------CREATE PLANTING AND HARVEST WINDOWS (BY SITE)------
```{r}
#| message: false
#| warning: false

library(daymetr)
library(slider)

site_calendar <- train_full %>%
  mutate(
    plant_doy = yday(date_planted),
    harvest_doy = yday(date_harvested)
  ) %>%
  group_by(site) %>%
  summarize(
    typical_plant_doy = round(median(plant_doy, na.rm = TRUE)),
    typical_harvest_doy = round(median(harvest_doy, na.rm = TRUE)),
    .groups = "drop"
  )
```

#-----------ADDING P/H WINDOWS TO TESTING AND TRAINING DATA------------
```{r}
#| message: false
#| warning: false

test_full <- test_full %>%
  left_join(site_calendar, by = "site") %>%
  mutate(
    plant_doy = typical_plant_doy,
    harvest_doy = typical_harvest_doy
  ) %>%
  mutate(
    plant_doy = coalesce(plant_doy, typical_plant_doy, global_plant_doy),
    harvest_doy = coalesce(harvest_doy, typical_harvest_doy, global_harvest_doy)
  ) %>%
  left_join(weather_features, by = c("site", "year"))
```

#-----FILLING IN MISSING DATE OF YEAR VALUES-------
```{r}
#| message: false
#| warning: false

global_plant_doy <- round(median(train_full$plant_doy, na.rm = TRUE))
global_harvest_doy <- round(median(train_full$harvest_doy, na.rm = TRUE))
```

#------ADDING MISSING VALUES TO TRAINING AND TESTING DATA-------
```{r}
#| message: false
#| warning: false

train_full <- train_full %>%
  mutate(
    plant_doy = coalesce(plant_doy, typical_plant_doy, global_plant_doy),
    harvest_doy = coalesce(harvest_doy, typical_harvest_doy, global_harvest_doy)
  )

test_full <- test_full %>%
  mutate(
    plant_doy = coalesce(plant_doy, typical_plant_doy, global_plant_doy),
    harvest_doy = coalesce(harvest_doy, typical_harvest_doy, global_harvest_doy)
  )
```

#-------CREATING A SITE-YEAR LIST-------
```{r}
#| message: false
#| warning: false

all_site_years <- bind_rows(
  train_full %>% distinct(site, year, latitude, longitude, plant_doy, harvest_doy),
  test_full %>% distinct(site, year, latitude, longitude, plant_doy, harvest_doy)
) %>%
  arrange(site, year) %>%
  mutate(
    longitude = if_else(longitude > 0 & longitude < 180, -longitude, longitude)
  )

head(all_site_years)
```

#---------FIXING ANY POS LONGITUDE NUMBERS (NORTH AMERICA HAS NEG LONGITUDE)--------
```{r}
#| message: false
#| warning: false

  all_site_years <- all_site_years %>%
   mutate(
     longitude = if_else(longitude > 0 & longitude < 180, -longitude, longitude)
   )
```

#----------PULLING DAYMET DATA---------- -->
```{r}
#| message: false
#| warning: false

# get_daymet_one <- function(site, year, lat, lon) {
#   tryCatch({
#     out <- download_daymet(
#       site = paste0(site, "_", year),
#       lat = lat,
#       lon = lon,
#       start = year,
#       end = year,
#       internal = TRUE,
#       silent = TRUE
#     )
# 
#     out$data %>%
#       as_tibble() %>%
#       mutate(
#         site = site,
#         year = year
#       )
#   }, error = function(e) {
#     message(paste("Skipping:", site, year, "lat =", lat, "lon =", lon))
#     return(NULL)
#   })
# }
# 
# weather_list <- purrr::pmap(
#   all_site_years,
#   ~ get_daymet_one(site = ..1, year = ..2, lat = ..3, lon = ..4)
# )
# 
# weather_daily <- bind_rows(weather_list)
# 
# 
# write_csv(weather_daily,
#           "../data/weather_daily.csv"
#           ) #saving pulled weather data to avoid repeated pulling
```

#------------FEATURE ENGINEERING WEATHER VARIABLES------------
```{r}
#| message: false
#| warning: false

weather_daily <- read_csv("../data/weather_daily.csv") # first load saved weather data

weather_daily <- weather_daily %>%
  mutate(
    tmean = (tmax..deg.c. + tmin..deg.c.) / 2,
    tmax_cap = pmin(tmax..deg.c., 30),
    tmin_cap = pmax(tmin..deg.c., 10),
    gdd10_30 = pmax(((tmax_cap + tmin_cap) / 2) - 10, 0),
    hot_day_32 = as.integer(tmax..deg.c. > 32),
    hot_night_21 = as.integer(tmin..deg.c. > 21),
    dry_day = as.integer(prcp..mm.day. < 1),
    rad_mj_m2 = (srad..W.m.2. * dayl..s.) / 1e6
  )
```

#-----------SUMMARIZE WEATHER VARIABLES------------
```{r}
#| message: false
#| warning: false

weather_features <- weather_daily %>%
  mutate(
    tmean = (tmax..deg.c. + tmin..deg.c.) / 2,
    gdd = pmax(tmean - 10, 0),
    hot_days = as.integer(tmax..deg.c. > 32),
    hot_nights = as.integer(tmin..deg.c. > 21),
    dry_day = as.integer(prcp..mm.day. < 1),
    rad_mj_m2 = (srad..W.m.2. * dayl..s.) / 1e6
  ) %>%
  group_by(site, year) %>%
  summarize(
    mean_temp = mean(tmean, na.rm = TRUE),
    max_temp = max(tmax..deg.c., na.rm = TRUE),
    min_temp = min(tmin..deg.c., na.rm = TRUE),
    total_prcp = sum(prcp..mm.day., na.rm = TRUE),
    rain_days = sum(prcp..mm.day. >= 1, na.rm = TRUE),
    hot_days = sum(hot_days, na.rm = TRUE),
    hot_nights = sum(hot_nights, na.rm = TRUE),
    gdd = sum(gdd, na.rm = TRUE),
    total_radiation = sum(rad_mj_m2, na.rm = TRUE),
    mean_vp = mean(vp..Pa., na.rm = TRUE),
    .groups = "drop"
  )
```

#-------ADDING WEATHER TO TRAINING AND TESTING DATA------- -->
```{r}
#| message: false
#| warning: false

train_full <- train_full %>%
  left_join(weather_features, by = c("site", "year"))

test_full <- test_full %>%
  left_join(weather_features, by = c("site", "year"))
```

#-----------SAVING THE FINISHED AND CLEAN DATA--------- -->
```{r}
#| message: false
#| warning: false

write_csv(train_full, "../data/train_full_weather.csv")
write_csv(test_full, "../data/test_full_weather.csv")
```


#--------LOADING FINISHED DATA SETS-------
```{r}
#| message: false
#| warning: false

train_full <- read_csv("../data/train_full_weather.csv", show_col_types = FALSE)
test_full  <- read_csv("../data/test_full_weather.csv", show_col_types = FALSE)
```

#----------CATEGORICAL TO NUMERICAL------------
```{r}
#| message: false
#| warning: false

train_full <- train_full %>%
  mutate(
    across(any_of(c("soilp_h", "om_pct", "soilk_ppm", "soilp_ppm")), as.numeric),
    across(any_of(c("grain_moisture")), as.numeric),
    across(any_of(c("date_planted", "date_harvested")), as.Date),
    site = as.character(site),
    hybrid = as.character(hybrid),
    previous_crop = as.character(previous_crop)
  )

test_full <- test_full %>%
  mutate(
    across(any_of(c("soilp_h", "om_pct", "soilk_ppm", "soilp_ppm")), as.numeric),
    across(any_of(c("grain_moisture")), as.numeric),
    across(any_of(c("date_planted", "date_harvested")), as.Date),
    site = as.character(site),
    hybrid = as.character(hybrid),
    previous_crop = as.character(previous_crop)
  )
```


#-------USING 2023 AS A TEST SET-------
```{r}
#| message: false
#| warning: false

train_dev <- train_full %>%
  filter(year < 2023) %>%
  select(-grain_moisture, -date_planted, -date_harvested) %>%
  mutate(
    site = as.character(site),
    hybrid = as.character(hybrid),
    previous_crop = as.character(previous_crop)
  )

test_2023 <- train_full %>%
  filter(year == 2023) %>%
  select(-grain_moisture, -date_planted, -date_harvested) %>%
  mutate(
    site = as.character(site),
    hybrid = as.character(hybrid),
    previous_crop = as.character(previous_crop)
  )

train_dev_knn <- train_dev %>%
  slice_sample(n = min(20000, nrow(train_dev)))
```

#---------CREATING THE RECIPES---------
```{r}
#| message: false
#| warning: false

library(recipes)
corn_recipe_xgb <- recipe(yield ~ ., data = train_dev) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_novel(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors())

knn_recipe <- recipe(yield ~ ., data = train_dev) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_novel(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())
```

#---------FITTING XGBOOST---------
```{r}
#| message: false
#| warning: false

library(tidymodels)
xgb_spec <- boost_tree(
  trees = 800,
  tree_depth = 6,
  learn_rate = 0.03,
  min_n = 5,
  loss_reduction = 0,
  sample_size = 0.8,
  mtry = 10
) %>%
  set_engine("xgboost") %>%
  parsnip::set_mode("regression")

xgb_wf <- workflow() %>%
  add_recipe(corn_recipe_xgb) %>%
  add_model(xgb_spec)

xgb_fit <- fit(xgb_wf, data = train_dev)
final_xgb_fit <- fit(xgb_wf, data = train_full)

xgb_pred_2023 <- predict(xgb_fit, new_data = test_2023) %>%
  bind_cols(test_2023 %>% 
              select(site, year, hybrid, yield))

xgb_metrics <- metrics(xgb_pred_2023, truth = yield, estimate = .pred)
xgb_metrics
```

#---------FITTING K-NEAREST NEIGHBORS---------
```{r}
library(kknn)

knn_spec <- nearest_neighbor(
  neighbors = 5,
  weight_func = "rectangular",
  dist_power = 2
) %>%
  set_engine("kknn") %>%
  set_mode("regression")

knn_wf <- workflow() %>%
  add_recipe(knn_recipe) %>%
  add_model(knn_spec)

knn_fit <- fit(knn_wf, data = train_dev_knn)

knn_pred_2023 <- predict(knn_fit, new_data = test_2023) %>%
  bind_cols(test_2023 %>% select(site, year, hybrid, yield))

knn_metrics <- metrics(knn_pred_2023, truth = yield, estimate = .pred)
knn_metrics
```

#------------PLOTTING XGBOOST------------
```{r}
xgb_rmse <- xgb_metrics %>% 
  filter(.metric == "rmse") %>% 
  pull(.estimate)
xgb_rsq  <- xgb_metrics %>% 
  filter(.metric == "rsq") %>% 
  pull(.estimate)

ggplot(xgb_pred_2023, aes(x = yield, y = .pred)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title = "XGBoost: Predicted vs Observed Yield",
    subtitle = paste0("RMSE = ", round(xgb_rmse, 2), " | R² = ", round(xgb_rsq, 3)),
    x = "Observed Yield",
    y = "Predicted Yield"
  ) +
  theme_minimal()
```

#------------PLOTTING K-NEAREST NEIGHBOR-----------
```{r}
knn_rmse <- knn_metrics %>% 
  filter(.metric == "rmse") %>% 
  pull(.estimate)
knn_rsq  <- knn_metrics %>% 
  filter(.metric == "rsq") %>% 
  pull(.estimate)

ggplot(knn_pred_2023, aes(x = yield, y = .pred)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title = "kNN: Predicted vs Observed Yield",
    subtitle = paste0("RMSE = ", round(knn_rmse, 2), " | R² = ", round(knn_rsq, 3)),
    x = "Observed Yield",
    y = "Predicted Yield"
  ) +
  theme_minimal()
```

#--------------COMPARING MODELS------------
```{r}
model_comparison <- bind_rows(
  xgb_metrics %>% 
    mutate(model = "XGBoost"),
  knn_metrics %>% 
    mutate(model = "K-Nearest Neighbor (kNN)")
)

print(model_comparison)
```

#-----------REFIT FINAL MODEL ON ALL TRAINING DATA
```{r}
#| message: false
#| warning: false

final_xgb_fit <- fit(xgb_wf, data = train_full)
final_knn_fit <- fit(knn_wf, data = train_full)
```

#------------PREDICT 2024-------------------
```{r}
xgb_pred_2024 <- predict(final_xgb_fit, new_data = test_full) %>%
  bind_cols(test_full %>% select(site, year, hybrid)) %>%
  rename(predicted_yield = .pred)

knn_pred_2024 <- predict(final_knn_fit, new_data = test_full) %>%
  bind_cols(test_full %>% select(site, year, hybrid))

head(xgb_pred_2024)
head(knn_pred_2024)
```

#--------------SAVE PREDICTIONS------------
```{r}
write_csv(xgb_pred_2024, "../output/xgb_predictions_2024.csv")
write_csv(knn_pred_2024, "../output/knn_predictions_2024.csv")
```
```{r}
test_full_final <- test_sub %>%
  left_join(test_meta, by = c("site", "year")) %>%
  left_join(test_soil, by = c("site", "year")) %>%
  left_join(weather_features, by = c("site", "year")) %>%
  left_join(site_calendar, by = "site") %>%
  mutate(
    plant_doy = typical_plant_doy,
    harvest_doy = typical_harvest_doy
  )
```


#-------------ADDING TO SUBMISSION FILE-------------
```{r}
xgb_pred_2024 <- predict(final_xgb_fit, new_data = test_full_final)

submission <- read_csv("../data/testing/testing_submission.csv", show_col_types = FALSE)

submission <- submission %>%
  mutate(yield_mg_ha = xgb_pred_2024$.pred)
```


#-------------VARIABLE IMPORTANCE------------
```{r}
final_xgb_fit %>%
  extract_fit_parsnip() %>%
  vip(num_features = 15)
```


```{r}
knitr::purl("finalproject_code.qmd", output = "finalproject_script.R", documentation = 0)
```

